openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2021-03-17 paper/policar2020-opentsne.tex
minor fixes

\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology are often used to reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t-SNE algorithm and its many
extensions.
The library is fast and enables users to embed data sets containing millions of
data points in a matter of minutes on consumer-grade laptop computers.
Unique to \opentsne\ is also the ability to map new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

@@

Despite its utility, t-SNE has often been criticized for its limited
scalability, lack of global organization -- t-SNE identifies well-defined clusters that
may be arbitrarily scattered throguhout the low-dimensional space -- and the absence of
theoretically-founded methods to map new data into existing
embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
shortcomings have recently been addressed. Linderman \textit{et al.} developed
an interpolation-based approximation scheme which massively improved the
scalability of t-SNE, achieving linear time complexity
in the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
several techniques to improve global cluster coherence, including estimating
similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our
previous work, we presented a principled approach for embedding new samples into
existing visualizations~\cite{policar2019embedding}.

@@

\opentsne\ is compatible with the Python data science ecosystem and
libraries including \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
providing a familiar and intuitive application program interface to new
users. Its modular design encourages extensibility and experimentation with
various settings and changes to the analysis pipeline. We have released
\opentsne\ to the open-source community to make recent theoretical advancements
more accessible to a wider audience. At
the time of the writing, the library averages thousands of weekly
downloads and has received almost 700 GitHub stars, an appreciation score gained
by only $0.3\%$ of public Python repositories. The source code is available at
\url{https://github.com/pavlin-policar/openTSNE} and supports installations from
\textsf{PyPI} and \textsf{conda-forge}, the two most widely adopted Python
package managers.

Accessibility of the latest theoretical advancments in t-SNE is one of the core
design principles of \opentsne. This range of advancments is easily accessible through an intuitive
programming interface, where we closely follow the style
\textsf{scikit-learn}~\cite{sklearn_api}, a popular and widely used
library for machine learning. The following snippet exemplifies this interface
and was used to generate the four visualizations in Fig.~\ref{fig:tasic}, demonstrating the
effect of a particular recent theoretical advancments in the t-SNE algorithm.

@@

Unlike other popular dimensionality reduction techniques such as principal
component analysis or autoencoders, t-SNE is a non-parametric method and does
not define an explicit mapping to the embedding space. Therefore embeddings of
new data points need to be found through
optimization~\cite{policar2019embedding}. \opentsne\ is currently the only
publicly available library allowing users to add new samples to existing
embeddings in a principled manner.

>>>>

\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology are often used to reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t\nobreakdash-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t\nobreakdash-SNE algorithm and its many
extensions.
The library is fast and enables users to embed data sets containing millions of
data points in a matter of minutes on consumer-grade laptop computers.
Unique to \opentsne\ is also the ability to map new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

@@

Despite its utility, t-SNE has often been criticized for its limited
scalability, lack of global organization -- t-SNE identifies well-defined clusters that
may be arbitrarily scattered throughout the low-dimensional space -- and the absence of
theoretically-founded methods to map new data into existing
embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
shortcomings have recently been addressed. Linderman \textit{et al.} developed
an interpolation-based approximation scheme which massively improved the
scalability of t-SNE, achieving linear time complexity
in the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
several techniques to improve global cluster coherence, including estimating
similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our
previous work, we presented a principled approach for embedding new samples into
existing visualizations~\cite{policar2019embedding}.

@@

\opentsne\ is compatible with the Python data science ecosystem and
libraries including \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
providing a familiar and intuitive application program interface to new
users. Its modular design encourages extensibility and experimentation with
various settings and changes to the analysis pipeline. We have released
\opentsne\ to the open-source community to make recent theoretical advancements
more accessible to a wider audience. At
the time of the writing, the library averages thousands of weekly
downloads and has received over 750 GitHub stars, an appreciation score gained
by only $0.3\%$ of public Python repositories. The source code is available at
\url{https://github.com/pavlin-policar/openTSNE} and supports installations from
\textsf{PyPI} and \textsf{conda-forge}, the two most widely adopted Python
package managers.

Accessibility of the latest theoretical advancements in t-SNE is one of the core
design principles of \opentsne. This range of advancements is easily accessible through an intuitive
programming interface, where we closely follow the style
\textsf{scikit-learn}~\cite{sklearn_api}, a popular and widely used
library for machine learning. The following snippet exemplifies this interface
and was used to generate the four visualizations in Fig.~\ref{fig:tasic}, demonstrating the
effect of a particular recent theoretical advancements in the t-SNE algorithm.

@@

Unlike other popular dimensionality reduction techniques such as principal
component analysis or autoencoders, t-SNE is a non-parametric method and does
not define an explicit mapping to the embedding space. Therefore, embeddings of
new data points need to be found through
optimization~\cite{policar2019embedding}. \opentsne\ is currently the only
publicly available library allowing users to add new samples to existing
embeddings in a principled manner.
